Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-30 draft.tex
.

Theoretical models of homeostasis are generally based on a form of feedback control. As such, they necessarily contain an error estimation. This could be based e.g. on the neuron's firing rate of the recent past measured in a given window, or, on a physiological level, e.g. a concentration of some molecule within some part of the cell structure. In any case, the error estimation requires setting some predefined target that the feedback control will attempt to achieve. This, of course, raises the question of what would be an appropriate choice of homeostatic targets.

@@

\subsection{Diffusive Homeostasis}
In this thesis, we examined \textit{diffusive homeostasis} as a possible candidate to replace the previously used model of intrinsic plasticity. The idea and modeling of diffusive homeostasis was adopted from a paper by Sweeney et al. \cite{Sweeney_Paper}, which models neural tissue as a two-dimensional surface and a set of points representing the neurons' positions. The group of excitatory neurons acted as a point-source of nitric oxide (NO), as well as as a sensor for the NO-concentration at each individual position. The production and sensing of NO forms the basis of a feedback loop: The individual NO-readout is fed into a comparator which causes an appropriate change within the internal firing threshold of the neuron, in turn altering the neuron's firing rate. The control system is then closed by linking the rate of NO-production to the neuron's firing rate. The results by Sweeney et al. suggest that diffusive signaling could resolve the dichotomy between overall stability of firing activity and the need to allow flexibility among individual neurons' activities. Intuitively speaking, through the diffusive signal, each neurons receives a "mixture" of its own activity and its neighboring neurons. Thereby individual tuning of firing rates is suppressed while the overall population activity is kept at a constant level.

@@

>>>>

Theoretical models of homeostasis are generally based on a form of feedback control. In control theory, this is also known as a \textit{closed loop control}. All these systems include a number of necessary functional compartments, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.
\newpage
\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
Generally, a closed loop control must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. However, it might not be possible to directly quantify this desired output in terms of some internal state of the process. For example, the firing rate of a neuron is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}. As a necessary next step, the (potentially transformed) output is fed into a \textit{comparator}, which, unsurprisingly, compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. However, the term \textit{homeostasis} generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. Additionally, this is also the point in the control loop, where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal, one can classify control loops by three types:
\newpage
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination.

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
When analyzing or modeling a biological system, a key step characterizing a control loop should be to identify the previously described compartments. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and "straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. As stated earlier, "firing rate" is an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a "sensible" quantity. In a nerve cell, this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible, and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section} and \ref{NO_Experiments_Section}. Given a representation of activity has been detected, its deviation from a reference value will cause other physiological compartments to affect the neuron's ability to generate action potentials.

  , they necessarily contain an error estimation. This could be based e.g. on the neuron's firing rate of the recent past measured in a given window, or, on a physiological level, e.g. a concentration of some molecule within some part of the cell structure. In any case, the error estimation requires setting some predefined target that the feedback control will attempt to achieve. This, of course, raises the question of what would be an appropriate choice of homeostatic targets.

@@

\subsection{Diffusive Homeostasis}\label{DIff_Hom_Intro_Section}
In this thesis, we examined \textit{diffusive homeostasis} as a possible candidate to replace the previously used model of intrinsic plasticity. The idea and modeling of diffusive homeostasis was adopted from a paper by Sweeney et al. \cite{Sweeney_Paper}, which models neural tissue as a two-dimensional surface and a set of points representing the neurons' positions. The group of excitatory neurons acted as a point-source of nitric oxide (NO), as well as as a sensor for the NO-concentration at each individual position. The production and sensing of NO forms the basis of a feedback loop: The individual NO-readout is fed into a comparator which causes an appropriate change within the internal firing threshold of the neuron, in turn altering the neuron's firing rate. The control system is then closed by linking the rate of NO-production to the neuron's firing rate. The results by Sweeney et al. suggest that diffusive signaling could resolve the dichotomy between overall stability of firing activity and the need to allow flexibility among individual neurons' activities. Intuitively speaking, through the diffusive signal, each neurons receives a "mixture" of its own activity and its neighboring neurons. Thereby individual tuning of firing rates is suppressed while the overall population activity is kept at a constant level.

@@

Generally, it has been argued that the set of rules used in these recurrent networks represent a realization of a minimum ensemble of functional compartments that are required in order to give rise to the desired capabilities. The network's ability of self-organization has not only been investigated in terms of network structure itself. It also successfully performed in unsupervised sequence-learning tasks, suggesting that the SORN can acquire and maintain associative memory \cite{Hartmann_2016}.
